---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tailr -- Tail recursion optimisations for R programming

[![Licence](https://img.shields.io/badge/licence-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-green.svg)](/commits/master)
[![packageversion](https://img.shields.io/badge/Package%20version-0.1.2.9000-green.svg?style=flat-square)](commits/master)
[![Travis build status](https://travis-ci.org/mailund/tailr.svg?branch=master)](https://travis-ci.org/mailund/tailr)
[![Appveyor build status](https://ci.appveyor.com/api/projects/status/1d36yh8klursko82/branch/master?svg=true)](https://ci.appveyor.com/project/mailund/tailr/branch/master)
[![Coverage status](https://codecov.io/gh/mailund/tailr/branch/master/graph/badge.svg)](https://codecov.io/github/mailund/tailr?branch=master)
[![Coverage status](http://coveralls.io/repos/github/mailund/tailr/badge.svg?branch=master)](https://coveralls.io/github/mailund/tailr?branch=master)
[![CRAN status](http://www.r-pkg.org/badges/version/tailr)](https://cran.r-project.org/package=tailr)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/grand-total/tailr)](https://cran.r-project.org/package=tailr)
[![minimal R version](https://img.shields.io/badge/R-%E2%89%A53.2-blue.svg)](https://cran.r-project.org/)


Recursive functions are the natural way to express iterations in a functional programming langauge, but in R, they can be significantly slower than loop-versions and for moderately long sequences or moderately deep trees, recursive functions will reach a limit imposted on them by the stack limit.

There are known solutions to these problems, as long as functions are written to be tail-recursive, meaning that the return value of a function is either a base value or another recursive call, but where we do not call recursively to then do something with the result.

The goal of `tailr` is to automatically transform tail-recursive functions into loops or trampolines.

## Installation

You can install the released version of `tailr` from CRAN using

``` r
install.packages("tailr")
```

You can install tailr from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mailund/tailr")
```


## Examples

Consider a classical recursive function, `factorial`:

```{r}
factorial <- function(n) {
    if (n <= 1) 1
    else n * factorial(n - 1)
}
```

(I know R already has a builtin factorial function, but please ignore that). This function will compute the factorial of `n`, but if `n` is too large, it will exceed the stack limit:

```r
> factorial(3000)
Error: C stack usage  7970184 is too close to the limit
```

A classical way out of this problem is to turn it into a tail-recursive function:

```{r}
factorial <- function(n, acc = 1) {
    if (n <= 1) acc
    else factorial(n - 1, acc * n)
}
```

R doesn't implement the tail-recursion optimisation, though, so it doesn't help us.

```r
> factorial(3000)
Error: C stack usage  7970184 is too close to the limit
```

With `tailr` we can, automatically, translate a tail-recursive function into a looping one, essentially implementing the tail-recursion optimisation this way.

```{r}
tr_factorial <- tailr::loop_transform(factorial, byte_compile = FALSE)
```

I have disabled byte compilation to make running time comparisons fair below; by default it is enabled. For a function as simple as `factorial`, though, byte compiling will not affect the running time in any substantial amount. 

This version, because it looks instead of recurse, doesn't have the stack limit problem:

```{r}
tr_factorial(3000)
```


We get the result `Inf` because the number we compute is too large to represent on the computer, but that is not the point of the example. The point is that the recursion doesn't get too deep for the stack because we avoid recursion alltogether.

With something as simple as computing the factorial, it is easy to write a looping function by hand, and it will be much faster than both the (tail-)recursive and the transformed function:

```{r}
loop_factorial <- function(n) {
    val <- 1
    while (n > 1) {
        val <- n * val
        n <- n - 1
    }
    val
}


n <- 1000
bm <- microbenchmark::microbenchmark(factorial(n), 
                                     tr_factorial(n), 
                                     loop_factorial(n))
bm
boxplot(bm)
```

The transformed version runs in about the same time as the recursive one, but the looping function is much faster.


However, consider a more complicated example. Using the `pmatch` package, we can create a linked list data structure as this:

```{r}
library(pmatch)
llist := NIL | CONS(car, cdr : llist)
```

A natural way to process linked lists using pattern matching is to write recursive functions that matches different patterns of their input. A function for computing the length of a linked list can look like this:

```{r}
llength <- case_func(acc = 0,
   NIL -> acc,
   CONS(car, cdr) -> llength(cdr, acc + 1)
)
```

```{r}
tr_llength <- tailr::loop_transform(llength)
```

The function we generate is rather complicated

```{r}
body(tr_llength)
```

but, then, it is not one we want to manually inspect in any case.

It is not too hard to implement this function with a loop either, but it is not as simple as the recursive function:

```{r}
is_nil <- case_func(NIL -> TRUE, otherwise -> FALSE)
loop_llength <- function(llist) {
    len <- 0
    while (!is_nil(llist)) {
        len <- len + 1
        llist <- llist$cdr
    }
    len
}
```

If we compare the running time for these three functions, the transformed function is faster than the recursive but not as fast as the iterative:

```{r}
make_llist <- function(n) {
    l <- NIL
    for (i in 1:n) {
        l <- CONS(i, l)
    }
    l
}
test_llist <- make_llist(100)
bm <- microbenchmark::microbenchmark(llength(test_llist),
                                     tr_llength(test_llist),
                                     loop_llength(test_llist))
bm
boxplot(bm)
```


More examples:

```{r}
llcontains <- case_func(x,
    NIL -> FALSE,
    CONS(car, cdr) -> if (car == x) TRUE else llcontains(cdr, x)
)

tr_llcontains <- tailr::loop_transform(llcontains)

loop_contains <- function(lst, x) {
    while (!is_nil(lst)) {
        if (x == lst$car) return(TRUE)
        else lst <- lst$cdr
    }
}

lst <- make_llist(100)
bm <- microbenchmark::microbenchmark(llcontains(lst, 1001),
                                     tr_llcontains(lst, 1001),
                                     loop_contains(lst, 1001))
bm
boxplot(bm)

```

```{r}
llrev <- pmatch::case_func(acc = NIL,
    NIL -> acc,
    CONS(car, cdr) -> llrev(cdr, CONS(car, acc))
)

bubble <- case_func(swapped = FALSE, acc = NIL,
    CONS(first, CONS(second, rest)) -> 
        if (first > second) bubble(CONS(first, rest), TRUE, CONS(second, acc))
        else bubble(CONS(second, rest), swapped, CONS(first, acc)),
    CONS(x, NIL) -> list(new_list = llrev(CONS(x, acc)), swapped = swapped)
)

bubble_sort <- function(lst) {
    if (is_nil(lst)) return(lst)
    bind[lst, swapped] <- bubble(lst)
    while (swapped) {
        bind[lst, swapped] <- bubble(lst)
    }
    lst
}

lst <- CONS(3, CONS(2, CONS(5, CONS(1, NIL))))
bubble_sort(lst)
```

```{r}
tr_llrev <- pmatch::case_func(acc = NIL,
    NIL -> acc,
    CONS(car, cdr) -> llrev(cdr, CONS(car, acc))
)

tr_llrev <- tailr::loop_transform(tr_llrev)

tr_bubble <- case_func(swapped = FALSE, acc = NIL,
    CONS(first, CONS(second, rest)) -> 
        if (first > second) tr_bubble(CONS(first, rest), TRUE, CONS(second, acc))
        else tr_bubble(CONS(second, rest), swapped, CONS(first, acc)),
    CONS(x, NIL) -> list(new_list = tr_llrev(CONS(x, acc)), swapped = swapped)
)
tr_bubble <- tailr::loop_transform(tr_bubble)

tr_bubble_sort <- function(lst) {
    if (is_nil(lst)) return(lst)
    bind[lst, swapped] <- tr_bubble(lst)
    while (swapped) {
        bind[lst, swapped] <- tr_bubble(lst)
    }
    lst
}

lst <- CONS(3, CONS(2, CONS(5, CONS(1, NIL))))
tr_bubble_sort(lst)
```


```{r}
loop_llrev <- function(lst) {
    acc <- NIL
    while (!is_nil(lst)) {
        acc <- CONS(lst$car, acc)
        lst <- lst$cdr
    }
    acc
}
loop_bubble <- function(lst, swapped = FALSE) {
    acc <- NIL
    repeat {
        if (is_nil(lst$cdr)) 
            return(list(new_list = loop_llrev(CONS(lst$car, acc)),
                        swapped = swapped))
        
        first <- lst$car
        second <- lst$cdr$car
        rest <- lst$cdr$cdr
        if (first > second) {
            acc <- CONS(second, acc)
            lst <- CONS(first, rest)
            swapped <- TRUE
        } else {
            acc <- CONS(first, acc)
            lst <- CONS(second, rest)
        }
    }
}

loop_bubble_sort <- function(lst) {
    if (is_nil(lst)) return(lst)
    bind[lst, swapped] <- loop_bubble(lst)
    while (swapped) {
        bind[lst, swapped] <- loop_bubble(lst)
    }
    lst
}

lst <- CONS(3, CONS(2, CONS(5, CONS(1, NIL))))
loop_bubble_sort(lst)
```

```{r}
lst <- make_llist(10)
bm <- microbenchmark::microbenchmark(bubble_sort(lst),
                                     tr_bubble_sort(lst),
                                     loop_bubble(lst))
bm
boxplot(bm)
```

The module primarily solves the problem of exceeding the stack space. The transformed functions are not as fast as those we can code by hand using loops. It *should* be possible to improve on the running time of the transformed functions, however, with some program analysis... This analysis should be included in the time usage analysis, though, which will probably still come out saying that manually programmed looping versions are faster than transformed functions. Recursive functions can be a lot easier to read, though, than their corresponding looping versions, especially with pattern matching.


