---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tailr -- Tail recursion optimisations for R programming

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-orange.svg)](/commits/master)
[![lifecycle](http://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Licence](https://img.shields.io/badge/licence-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)

[![Travis build status](https://travis-ci.org/mailund/tailr.svg?branch=master)](https://travis-ci.org/mailund/tailr)
[![Appveyor build status](https://ci.appveyor.com/api/projects/status/1d36yh8klursko82/branch/master?svg=true)](https://ci.appveyor.com/project/mailund/tailr/branch/master)
[![Coverage status](https://codecov.io/gh/mailund/tailr/branch/master/graph/badge.svg)](https://codecov.io/github/mailund/tailr?branch=master)
[![Coverage status](http://coveralls.io/repos/github/mailund/tailr/badge.svg?branch=master)](https://coveralls.io/github/mailund/tailr?branch=master)

[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.1-blue.svg)](https://cran.r-project.org/)
[![packageversion](https://img.shields.io/badge/Package%20version-0.0.0.9000-orange.svg?style=flat-square)](commits/master)
<!--[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/tailr)](https://cran.r-project.org/package=tailr)
-->
 
---
 


The goal of tailr is to automatically transform tail-recursive functions into loops or trampolines.

## Installation

You can install tailr from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mailund/tailr")
```


## Examples

We can take a classical recursive function and write it in a tail-recursive form using an accumulator:

```{r}
factorial <- function(n, acc = 1) {
    if (n <= 1) { acc }
    else { factorial(n - 1, acc * n) }
}
```

We can then, automatically, translate that into a looping version:

```{r}
tr_loop_factorial <- tailr::loop_transform(factorial)
tr_loop_factorial
```

We can then compare the running time with the recursive function and a version that is written using a loop:

```{r}
loop_factorial <- function(n) {
    val <- 1
    while (n > 1) {
        val <- n * val
        n <- n - 1
    }
    val
}


n <- 100
microbenchmark::microbenchmark(factorial(n), loop_factorial(n), tr_loop_factorial(n))
```

There is some overhead in the translated version. I believe this is the parallel assignments that replace the recursive call, see Issue #7.
